Type theories as quotient inductive-recursive types

### 2025-02-11 tex/main.tex
Section 3.4.1

\begin{abstract}
  Defining type theory in type theory is a long-standing benchmark for the strength and expressivity of type theory and a proof assistant.
  Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory in intensional type theory, with the price to be paid that terms frequently contain transports along equality constructors.
  As a result, proofs need to painstakingly account for coercions alongside the interesting part.
  This %mess
  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation.
  In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
  As a case study, we investigate type theory with a universe and $\Pi$-types, and compare our definition with quotient inductive-inductive definitions.
\end{abstract}

@@

\subsection{Quotient inductive-inductive types}
% \subsection{Quotient inductive(-inductive)-recursive types}

@@

Finally, we show that these simultaneously defined functions are coherent.
\begin{proposition}[Coherence]
  For every $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the following identities
  \danger
  \[
    \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
    \qquad\text{and}\qquad
    [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
  \]
  hold for any type $A$ and any term $t$.
\end{proposition}
\begin{proof}
  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
\Fredrik{This heterogeneous equality has not been used so far; it would be good to not have to introduce it all, except perhaps locally in the proof here.}
\LT{How about deferring introducing heterogeneous equality until \cref{sec:formal-implementation}?}
\end{proof}

\subsubsection{Multiple lifting and the coherence of type substitution}
\Fredrik{This section gets rather technical. Keeping the page limit in mind, I wonder if we should only briefly summarise it in the main text?}
\LT{Move this to the appendix?}

For the coherence of~$[\blank]_{\Ty}\blank$, we have to generalise the desired identity to account for the lifting, since a naive induction does not work: for example, unfolding $[\sigma]_{\Ty}\,(\Pi\;A\;B)$ leads to
\[
  \Pi\;([\sigma]_{\Ty}\,A)\;([\sigma\uparrow A]_{\Ty}\,B)
\]
where the induction hypothesis $[\sigma]_{\Ty}\,B = [\emptysub]_{\Ty}\,B$ cannot be applied.
Instead, we will define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, which intuitively means multiple applications of lifting: $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
Then, we show the identity
$[\sigma \mathop{\upuparrows} \Xi ]_{\Ty}\,A = [\tau \mathop{\upuparrows} \Xi ]_{\Ty}\,A$
for arbitrary $\Xi$, subsuming the special case $[\sigma ]_{\Ty}\,A = [\tau ]_{\Ty}\,A$ for $\Xi$ being empty.

First, define an inductive-recursive type of \emph{telescopes} under a context $\Gamma$ simultaneously with a concatenation $\blank \dplus \blank$ as follows.

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \Tel               & :                   && \Ctx \to \Set                                     \\
  \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
  \Gamma & \dplus \emptytel && \reduce \Gamma \\
  \Gamma & \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
Then, we define the multiple lifting $\blank \upuparrows \blank$.
Observe that its type, generalising the type \eqref{eq:type-of-lifting} of a single lifting $\blank\uparrow\blank$, should be
\[
\blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
\to \Sub\;(\Gamma \dplus ([\sigma]_{\Tel}\,\Xi))\;(\Delta \dplus \Xi)
\]
asking for telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$, but $[\blank]_{\Tel}\blank$ also needs the multiple lifting for the case $[ \sigma ]_{\Tel}\,(\Xi, A)$ (why?).
Hence, we arrive the conclusion that telescope substitution and the multiple lifting by a telescope need to be defined mutually:
\begin{minipage}{.5\linewidth}
\begin{alignat*}{5}
  & [ \sigma ]_{\Tel}\,\emptytel && \reduce \emptytel \\
  & [ \sigma ]_{\Tel}\,(\Xi, A)  && \reduce [ \sigma ]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\linewidth}
\begin{alignat*}{5}
  & \sigma \upuparrows \emptytel && \reduce \sigma \\
  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip

To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows [\tau]_{\Tel} \Xi); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.

Instead of resorting to \transp, we also turn telescope substitution $[\blank]_{\Tel}\blank$ and the multiple lifting into definitions by overlapping patterns by adding following clauses:

@@

Now, we proceed with the coherence property for the recursive functions introduced so far.
The first set of equality constructors are easy to prove, since the lifting and substitution operations compute independently for these cases:
\begin{proposition}[Coherence, I]
  \danger
  For the following equality constructors $p : \sigma = \tau$
  \[
    \mathsf{idr} : \sigma;\idS = \sigma, \quad
    \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma), 
  \]
  the identities $[\sigma]_{\Tel}\,\Xi \judgeq [\tau]_{\Tel}\,\Xi$ and $[\sigma ]_{\Ty}\,A \judgeq [\tau ]_{\Ty}\,A$
  hold strictly for any telescope $\Xi$ and any type $A$ and
  \[
    \sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}
    \tau   \upuparrows \Xi
  \]
  holds propositionally for any telescope $\Xi$.
\end{proposition}
\begin{proof}
  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds strictly by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
  Take $\mathsf{idr}$ as an example, we can see 
  \begin{align*}
    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows [\idS]_{\Tel} \Xi) ; (\idS \upuparrows \Xi) \\
                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
                                   & = \sigma \upuparrows \Xi
  \end{align*}
  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.

@@

We conclude this extension with $\Pi$-types by noting that single term substitution
\begin{alignat*}{3}
  \left<\blank\right> & : \implicit{\Gamma, A} \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
  \left< t \right>    & \defeq (\idS , t),
\end{alignat*}
used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$, 
can be introduced without any transport, which is $(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})$ in its QII definition.
Similarly, the structural rule for substitution on $\mathsf{app}$ can be derived
\begin{align*}
  [\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t) & = \mathsf{app}\,\lambda([\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t)) && \text{by $\Pi\beta^{-1}$} \\
                                              & = \mathsf{app}\,\left([\sigma]\,\lambda\,(\mathsf{app}\,t)\right) && \text{by $[]\lambda^{-1}$} \\
                                              & = \mathsf{app}\,([\sigma]\,t) && \text{by $\Pi\eta^{-1}$.}
\end{align*}
without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become strict.

@@

>>>>

\begin{abstract}
  Defining type theory in type theory is a long-standing benchmark for the strength and expressivity of type theory and proof assistants.
  Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory in intensional type theory, with the price to be paid that terms frequently contain transports along equality constructors.
  As a result, proofs need to painstakingly account for coercions alongside the interesting part.
  This %mess
  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation.
  In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
  As a case study, we investigate type theory with a universe and $\Pi$-types, and compare our development with one using quotient inductive-inductive types.
\end{abstract}

@@

\subsection{Quotient inductive-inductive and quotient inductive-recursive types}
% \subsection{Quotient inductive(-inductive)-recursive types}

@@

Finally, we show that these simultaneously defined functions are coherent.
\begin{proposition}[Coherence]
  For every constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, we have
  \danger
  \[
    \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
    \qquad\text{and}\qquad
    [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
  \]
  for any type $A$ and any term $t$.
\end{proposition}
\begin{proof}
  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
\Fredrik{This heterogeneous equality has not been used so far; it would be good to not have to introduce it all, except perhaps locally in the proof here.}
\LT{How about deferring introducing heterogeneous equality until \cref{sec:formal-implementation}?}
\end{proof}

\subsubsection{Multiple lifting and the coherence of type substitution}

We also need to prove the coherence for type substitution, i.e., that $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
To do this, we need to generalise the desired identity to account for liftings, since a naive induction does not work: for example, on $\Pi$-types, we have
\begin{align*}
  [\sigma]_{\Ty}\,(\Pi\;A\;B) & = \Pi\;([\sigma]_{\Ty}\,A)\;([\sigma\uparrow A]_{\Ty}\,B) \\
  [\tau]_{\Ty}\,(\Pi\;A\;B) & = \Pi\;([\tau]_{\Ty}\,A)\;([\tau\uparrow A]_{\Ty}\,B)
\end{align*}
but the induction hypothesis does not give us $[\sigma\uparrow A]_{\Ty}\,B = [\tau\uparrow A]_{\Ty}\,B$.
Instead, we will define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, which intuitively means multiple applications of lifting: $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
Then, for each constructor $p : \sigma = \tau$, we show the identity
$[\sigma \mathop{\upuparrows} \Xi ]_{\Ty}\,A = [\tau \mathop{\upuparrows} \Xi ]_{\Ty}\,A$
for arbitrary $\Xi$, subsuming the special case $[\sigma ]_{\Ty}\,A = [\tau ]_{\Ty}\,A$ when $\Xi$ is empty.

The type of telescopes over a ``base'' context $\Gamma$ $\Tel : \Ctx \to \Set$ are defined inductive-recursively simultaneously with a concatenation $\blank \dplus \blank$ as follows.

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \Tel               & :                   && \Ctx \to \Set                                     \\
  \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
  \Gamma & \dplus \emptytel && \reduce \Gamma \\
  \Gamma & \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
Equipped with telescopes, we can now generalise single lifting with type \eqref{eq:type-of-lifting} to multiple lifting with type
\[
\blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
\to \Sub\;(\Gamma \dplus ([\sigma]_{\Tel}\,\Xi))\;(\Delta \dplus \Xi)
\]
which requires us to also simultaneously define telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$.

\begin{minipage}{.5\linewidth}
\begin{alignat*}{5}
  & [ \sigma ]_{\Tel}\,\emptytel && \reduce \emptytel \\
  & [ \sigma ]_{\Tel}\,(\Xi, A)  && \reduce [ \sigma ]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\linewidth}
\begin{alignat*}{5}
  & \sigma \upuparrows \emptytel && \reduce \sigma \\
  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip

To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows [\tau]_{\Tel} \Xi); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
%
Instead of resorting to \transp, we also turn telescope substitution $[\blank]_{\Tel}\blank$ and the multiple lifting into definitions by overlapping patterns by adding the following clauses:

@@

Of course, we also have to prove coherence for the newly introduced recursive functions to verify that they are well defined.
\begin{proposition}[{Coherence for $[\blank]_{\Tel}$ and $\blank\upuparrows\blank$}]
  \danger
  For any equality constructor $p : \sigma = \tau$, we have
  $[\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$
  for any telescope $\Xi$.
\end{proposition}

As promised, by instantiating $\Xi$ to be empty, we also get coherence for $[\blank]_{\Ty}$.

@@

We conclude this extension with $\Pi$-types by noting that for $t : \Tm\,\Gamma\,(\Pi\,A\,B)$ and $u : \Tm\,\Gamma\,A$, also ``ordinary'' application
$t \mathop{\$} u \defeq [ (\idS , t) ]_{\Tm} (\mathsf{app}\,t)$ of type $\Tm\,\Gamma\,([(\idS , t)] B)$ can be defined without any transport. This is in contrast to the pure QII situation, where application is defined as $[ (\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}}) ]_{\Tm} (\mathsf{app}\,t)$.
%
%\begin{alignat*}{3}
%  \left<\blank\right> & : \implicit{\Gamma, A} \Tm\;\Gamma\;A \to \Sub\;\Gamma\;(\Gamma, A) \\
%  \left< t \right>    & \defeq (\idS , t),
%\end{alignat*}
%used to define the ordinal application: $t \mathop{\$} u \defeq [ \left< u \right> ]_{\Tm} (\mathsf{app}\,t)$, 
%can be introduced without any transport, which is $(\idS, \alert{\transfib{\Tm\;\Gamma}{[id]_T^{-1}}{\color{black}t}})$ in its QII definition.
Similarly, the structural rule for substitution on $\mathsf{app}$
\begin{align*}
  [\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t) & = \mathsf{app}\,\lambda([\sigma \uparrow A ]_{\Tm}(\mathsf{app}\;t)) && \text{by $\Pi\beta^{-1}$} \\
                                              & = \mathsf{app}\,\left([\sigma]\,\lambda\,(\mathsf{app}\,t)\right) && \text{by $[]\lambda^{-1}$} \\
                                              & = \mathsf{app}\,([\sigma]\,t) && \text{by $\Pi\eta^{-1}$.}
\end{align*}
can be derived without any transport (cf.\ \cite{Altenkirch2016a}), since \eqref{eq:def-type-subst-8} has become strict.

@@

\section{Omitted proofs}

\Fredrik{Do properly with restate-thm etc}
The first set of equality constructors are easy to prove, since the lifting and substitution operations compute independently for these cases:
\begin{proposition}[Coherence, I]
  \danger
  For the following equality constructors $p : \sigma = \tau$
  \[
    \mathsf{idr} : \sigma;\idS = \sigma, \quad
    \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma), 
  \]
  the identities $[\sigma]_{\Tel}\,\Xi \judgeq [\tau]_{\Tel}\,\Xi$ and $[\sigma ]_{\Ty}\,A \judgeq [\tau ]_{\Ty}\,A$
  hold strictly for any telescope $\Xi$ and any type $A$ and
  \[
    \sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}
    \tau   \upuparrows \Xi
  \]
  holds propositionally for any telescope $\Xi$.
\end{proposition}
\begin{proof}
  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds strictly by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
  Take $\mathsf{idr}$ as an example, we can see 
  \begin{align*}
    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows [\idS]_{\Tel} \Xi) ; (\idS \upuparrows \Xi) \\
                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
                                   & = \sigma \upuparrows \Xi
  \end{align*}
  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.
